The unphysicality of Hilbert spaces

### 2024-10-30 Articles/2023-HilbertQuantum/Submissions/Quantum Studies/HilbertQuantum.tex
HilbertQuantum: adding extra section for motivation

@@

Let us first clarify the standard we believe should be used to judge the physicality of a formal definition.\cite{redei_tension_2020, north_physics_2021} Suppose we have a physical system under study in a given condition. A mathematical definition is \textbf{physical} if it properly characterizes the physical system. That is, if it can be shown that the mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart.

For example, suppose we have a set of physically distinguishable elements (e.g. states, bodies, regions of space), meaning that we can tell any two apart through an experimental test. Experimentally, the best we can do is collect finite information in finite time. If given an arbitrarily long time, we can collect an arbitrarily large but finite amount of information. An element of the set, then, must be identified by, at most, a countably infinite sequence of ones and zeros so that, at some point, we can tell the difference between the two sequences and, therefore, the elements. The set of countable sequences of zeros and ones can be understood as the binary expansion of real numbers between zero and one, which has the cardinality of the continuum. Therefore, any mathematical definition requiring a set of physical elements to have cardinality greater than the continuum is \textbf{unphysical}. For example, non-separable Hilbert spaces have cardinality greater than the continuum and, therefore, are unphysical.\cite{pittphilsci18363}

Naturally, this does not mean all sets with a cardinality of the continuum or lower are physical. In general, testing physicality is harder as we need to check that all the mathematical features map to physical features and that no physical feature is left behind. Clearly, postulating that ``the state space of quantum states has cardinality no greater than that of the continuum'' is certainly \textbf{physical}, but this tells us little. Let us then see what features of Hilbert spaces we should consider physical in the context of quantum mechanics.

>>>>

\section{On physicality}

When we first drafted this paper, we naively thought that the advantage of having a well-posed mathematical structure that represents, and only represents, physically well-defined objects would be self-evident. It does not seem to be the case: a previous reviewer stated that ``the question of whether the limiting states are actually physical is rather pointless;'' conversely, others see no problem in working with QFT path integrals that are not mathematically well-defined. We are therefore in the absurd, from our perspective, position of having to defend that, yes, the math we need to use in physics has be mathematically precise and, yes, the math we need to use in physics has to represent, perhaps idealized, physically realizable objects. The cost of this failure is all around us: an ever increasing number of interpretations of quantum mechanics, abstract mathematical work whose connection to physics is unclear, theoretical physics more and more disconnected from experimental physics. The reader that, like us, finds this obvious can skip this section.

One problem is that most physicists see mathematics as just a tool for calculation, much like spectroscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say ``states are rays in a Hilbert space'' we are giving a physical specification. We are saying that there are objects in the real world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

To make this determination, we need to provide a standard to judge the physicality of a formal definition.\cite{redei_tension_2020, north_physics_2021} Suppose we are studying a physical system. A mathematical definition is \textbf{physical} if it properly characterizes the physical system in the given conditions. That is, if we are able to \textbf{justify on physical grounds} that the given mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart. The question of this paper, then, is whether the use of Hilbert spaces to characterize quantum systems can be justified on physical grounds, from physical requirements. This is the task at hand.

Some mathematicians were disappointed that we didn't discuss in depth approaches like rigged Hilbert spaces or the algebraic approach to quantum mechanics. Do rigged Hilbert spaces spell out exactly what each element of the Gelfand triple describes and give a physical justification that tells us exactly which of those are the physically correct specifications? Do algebraic approaches give a physical justification as to why things we measure in a lab must be modeled by a $C^*$-algebra, and consequently allow us to determine whether the use of Hilbert spaces to model states can be given (or not given) a physical justification? To our knowledge, and to the knowledge of the mathematicians that approached us, they do not. This does not mean that these works are of no interest to a physical audience. It just means they are not going to help with the task at hand.

While it seems obvious to us, it may be worth stating that \textbf{it is not up to mathematicians to justify the use of a particular mathematical structure in physics}. Ultimately, the justification cannot be solely in terms of mathematical matters, but rather it has to rely on that elusive ``physical intuition.'' In fact, the task is exactly to make that elusive ``physical intuition'' precise enough to turn it into mathematical definitions. This is the job of a physicist. In our specific case, the core argument we will use is that states described by measurable quantities with infinite or ill-defined expectation values are not physically realizable. To us, and to every experimental physicist we talked to, ``it doesn't make sense experimentally,'' and this is the best argument possible to show that something is unphysical. ``It can't be done.'' No argument is better. However, more than one previous reviewer found no problem with infinite (or undefined) expectation values, as they adopt an ``ensemble interpretation,'' and tried to convince us that we shouldn't either. How can progress on the mathematical foundations of quantum mechanics be made if an argument that is accepted by the vast majority of physicists is rejected?

Obviously, while the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implication for both the physics and the math. Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not include typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the original premise, the evidence-based requirement of physics, justifies the mathematical requirements.

To summarize, we are not interested in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. The fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, \textbf{physical} entities we prepare in a lab. Something we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?

@@

As we have discussed in the previous section, to argue for the physicality of a mathematical definition we need to make a justification \textbf{on physical grounds} that the definition captures a physical property.  We will see that the inner product space part of the Hilbert space definition is physical as it is required to to compute entropy and probabilities. We have also seen that to argue for the unphysicality of a mathematical definition we need to show that the definition implies objects that are physically unrealizable. We will see that the completeness part of the Hilbert space definition is unphysical as it forces us to include states with infinite (or undefined) expectation values.
